---
title: "Build and deploy a stroke prediction model using R"
date: "`r Sys.Date()`"
output: html_document
author: "Nikhil Padman"
---

# About Data Analysis Report

This RMarkdown file contains the report of the data analysis done for the project on building and deploying a stroke prediction model in R. It contains analysis such as data exploration, summary statistics and building the prediction models. The final report was completed on `r date()`. 

**Data Description:**

According to the World Health Organization (WHO) stroke is the 2nd leading cause of death globally, responsible for approximately 11% of total deaths.

This data set is used to predict whether a patient is likely to get stroke based on the input parameters like gender, age, various diseases, and smoking status. Each row in the data provides relevant information about the patient.


# Task One: Import data and data preprocessing

## Load data and install packages

```{r}
# Install Package
install.packages("pROC")
install.packages("tidyverse")
install.packages("caret")
install.packages("corrplot")
install.packages("mice")
install.packages("randomForest")

# Load required libraries
library(tidyverse)
library(caret)
library(dplyr)
library(tidyr)
library(ggplot2)
library(corrplot)
library(mice)
library(randomForest)
library(pROC)

getwd()

# Read the dataset
# Check if file exists first
if (file.exists("healthcare-dataset-stroke-data.csv")) {
  stroke_data <- read.csv("healthcare-dataset-stroke-data.csv")
  print("Data loaded successfully")
} else {
  print("File not found in current directory")
}

```


```{r}
# Display the first few rows and basic information
head(stroke_data)
str(stroke_data)
summary(stroke_data)
```


## Describe and explore the data

```{r}
# Basic data exploration
# Convert categorical variables to factors
stroke_data$gender <- as.factor(stroke_data$gender)
stroke_data$ever_married <- as.factor(stroke_data$ever_married)
stroke_data$work_type <- as.factor(stroke_data$work_type)
stroke_data$Residence_type <- as.factor(stroke_data$Residence_type)
stroke_data$smoking_status <- as.factor(stroke_data$smoking_status)
stroke_data$bmi <- as.numeric(stroke_data$bmi)
stroke_data$stroke <- as.factor(stroke_data$stroke)

# Get distribution of stroke cases
stroke_distribution <- stroke_data %>%
  group_by(stroke) %>%
  summarise(count = n(),
            percentage = n()/nrow(stroke_data)*100,
            .groups = 'drop')

print("Distribution of Stroke Cases:")
print(stroke_distribution)

# Calculate summary statistics by stroke status
stroke_summary <- stroke_data %>%
  group_by(stroke) %>%
  summarise(
    avg_age = mean(age),
    avg_glucose = mean(avg_glucose_level),
    avg_bmi = mean(bmi, na.rm = TRUE),
    hypertension_rate = mean(hypertension) * 100,
    heart_disease_rate = mean(heart_disease) * 100,
    .groups = 'drop'
  )

print("\nSummary Statistics by Stroke Status:")
print(stroke_summary)

# Create age groups for better visualization
stroke_data$age_group <- cut(stroke_data$age, 
                            breaks = c(0, 20, 40, 60, 80, 100),
                            labels = c("0-20", "21-40", "41-60", "61-80", "80+"))

# Age distribution analysis
age_distribution <- stroke_data %>%
  group_by(age_group, stroke) %>%
  summarise(count = n(),
            .groups = 'drop') %>%
  pivot_wider(names_from = stroke, 
              values_from = count,
              names_prefix = "stroke_")

print("\nAge Distribution by Stroke Status:")
print(age_distribution)

# Check missing values
missing_values <- colSums(is.na(stroke_data))
print("\nMissing Values in Each Column:")
print(missing_values)
```

```{r}
```


```{r}
# Create multiple visualizations
library(gridExtra)

# 1. Age Distribution Plot
age_plot <- ggplot(stroke_data, aes(x = age, fill = stroke)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  labs(title = "Age Distribution by Stroke Status",
       x = "Age",
       y = "Count") +
  theme_minimal()

# 2. BMI vs Glucose Level Plot
bmi_glucose_plot <- ggplot(stroke_data, aes(x = bmi, y = avg_glucose_level, color = stroke)) +
  geom_point(alpha = 0.5) +
  labs(title = "BMI vs Average Glucose Level",
       x = "BMI",
       y = "Average Glucose Level") +
  theme_minimal()

# 3. Stroke Distribution by Gender
gender_plot <- ggplot(stroke_data, aes(x = gender, fill = stroke)) +
  geom_bar(position = "fill") +
  labs(title = "Stroke Distribution by Gender",
       x = "Gender",
       y = "Proportion") +
  theme_minimal()

# 4. Medical Conditions Impact
medical_data <- stroke_data %>%
  gather(key = "condition", value = "status", 
         c("hypertension", "heart_disease")) %>%
  mutate(status = as.factor(status))

medical_plot <- ggplot(medical_data, aes(x = condition, fill = stroke)) +
  geom_bar(position = "fill") +
  labs(title = "Stroke Distribution by Medical Conditions",
       x = "Condition",
       y = "Proportion") +
  theme_minimal()

# Arrange all plots in a grid
grid.arrange(age_plot, bmi_glucose_plot, gender_plot, medical_plot, ncol = 2)

# Create summary statistics table
summary_stats <- data.frame(
  Metric = c("Total Observations", 
             "Missing BMI Values", 
             "Age Range",
             "Average Glucose Level Range"),
  Value = c(nrow(stroke_data),
            sum(is.na(stroke_data$bmi)),
            paste(min(stroke_data$age), "-", max(stroke_data$age)),
            paste(round(min(stroke_data$avg_glucose_level),2), "-", 
                  round(max(stroke_data$avg_glucose_level),2)))
)

print("Summary Statistics:")
print(summary_stats)
```


```{r}
# 1. First, let's handle missing values in BMI
# Create a copy of the dataset
stroke_processed <- stroke_data

# Impute missing BMI values using median by age group and gender
stroke_processed$bmi <- as.numeric(stroke_processed$bmi)  # Ensure BMI is numeric
stroke_processed <- stroke_processed %>%
  group_by(age_group, gender) %>%
  mutate(bmi = ifelse(is.na(bmi), median(bmi, na.rm = TRUE), bmi)) %>%
  ungroup()

# Check if any missing values remain
print("Remaining missing values after imputation:")
print(colSums(is.na(stroke_processed)))

# 2. Create derived features
stroke_processed <- stroke_processed %>%
  mutate(
    # BMI Category
    bmi_category = case_when(
      bmi < 18.5 ~ "Underweight",
      bmi >= 18.5 & bmi < 25 ~ "Normal",
      bmi >= 25 & bmi < 30 ~ "Overweight",
      bmi >= 30 ~ "Obese"
    ),
    
    # Glucose Category
    glucose_category = case_when(
      avg_glucose_level < 70 ~ "Low",
      avg_glucose_level >= 70 & avg_glucose_level < 100 ~ "Normal",
      avg_glucose_level >= 100 & avg_glucose_level < 126 ~ "Pre-diabetes",
      avg_glucose_level >= 126 ~ "Diabetes"
    ),
    
    # Combined health risk score
    health_risk_score = hypertension + heart_disease,
    
    # Age categories (already created, but let's make it more specific)
    age_category = case_when(
      age < 13 ~ "Child",
      age >= 13 & age < 20 ~ "Teen",
      age >= 20 & age < 40 ~ "Young Adult",
      age >= 40 & age < 60 ~ "Middle Aged",
      age >= 60 ~ "Senior"
    )
  )

# 3. Convert categorical variables to factors
categorical_vars <- c("bmi_category", "glucose_category", "age_category")
stroke_processed[categorical_vars] <- lapply(stroke_processed[categorical_vars], as.factor)

# 4. Create interaction features
stroke_processed <- stroke_processed %>%
  mutate(
    hypertension_heart = interaction(hypertension, heart_disease),
    age_hypertension = interaction(age_category, hypertension)
  )

# 5. Print summary of new features
print("\nSummary of new derived features:")
summary(stroke_processed[c("bmi_category", "glucose_category", "health_risk_score", "age_category")])

# 6. Verify the structure of processed dataset
str(stroke_processed)

# 7. Create a correlation matrix for numeric variables
numeric_vars <- stroke_processed %>%
  select_if(is.numeric) %>%
  select(-id)  # Remove ID column as it's not relevant for correlation

correlation_matrix <- cor(numeric_vars, use = "complete.obs")
print("\nCorrelation matrix of numeric variables:")
print(round(correlation_matrix, 2))

# 8. Save processed dataset
processed_data <- stroke_processed %>%
  select(-id)  # Remove ID column as it's not needed for modeling

# Display the first few rows of the processed dataset
print("\nFirst few rows of processed dataset:")
head(processed_data)
```


# Task Two: Build prediction models

```{r}
# 1. First, let's check and clean the stroke variable
print("Initial stroke value counts:")
table(stroke_data$stroke)

# 2. Clean and prepare the data
stroke_clean <- stroke_data %>%
  # Remove any rows with NA values
  na.omit() %>%
  # Ensure stroke is binary (0 or 1)
  filter(stroke %in% c(0, 1)) %>%
  # Convert stroke to factor with proper labels
  mutate(
    stroke = factor(stroke, 
                   levels = c(0, 1), 
                   labels = c("No_Stroke", "Stroke"))
  )

# 3. Verify the cleaning
print("\nCleaned stroke value counts:")
table(stroke_clean$stroke)

# 4. Create the model with clean data
# Split the data
set.seed(123)
split_index <- createDataPartition(stroke_clean$stroke, p = 0.8, list = FALSE)
train_data <- stroke_clean[split_index, ]
test_data <- stroke_clean[-split_index, ]

# Set up cross-validation
ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  sampling = "smote"
)

# Select features for modeling
features <- c("age", "gender", "hypertension", "heart_disease", 
             "ever_married", "work_type", "Residence_type", 
             "avg_glucose_level", "bmi", "smoking_status")

# Train the model
model_formula <- as.formula("stroke ~ age + gender + hypertension + heart_disease + 
                           ever_married + work_type + Residence_type + 
                           avg_glucose_level + bmi + smoking_status")

rf_model <- train(
  model_formula,
  data = train_data,
  method = "rf",
  metric = "ROC",
  trControl = ctrl,
  na.action = na.omit
)

# Make predictions
rf_pred <- predict(rf_model, test_data)
rf_pred_prob <- predict(rf_model, test_data, type = "prob")

# Calculate performance metrics
conf_matrix <- confusionMatrix(rf_pred, test_data$stroke)

# Print results
print("\nModel Performance:")
print(conf_matrix)

# Calculate and plot ROC curve
roc_obj <- roc(test_data$stroke, rf_pred_prob[,"Stroke"])
print(paste("\nAUC-ROC:", auc(roc_obj)))

# Plot ROC curve
plot(roc_obj, main = "ROC Curve for Stroke Prediction")

# Feature importance
importance <- varImp(rf_model)
print("\nFeature Importance:")
print(importance)

# Save the model
saveRDS(rf_model, "stroke_prediction_model.rds")
```




# Task Three: Evaluate and select prediction models

```{r}
# Enhance ROC curve visualization
plot(roc_obj,
     main = "ROC Curve for Stroke Prediction",
     col = "blue",
     lwd = 2,
     print.auc = TRUE,
     print.thres = TRUE,
     auc.polygon = TRUE,
     grid = TRUE,
     legacy.axes = TRUE)

# Add confidence intervals
ci.roc <- ci(roc_obj)
plot(ci.roc, col = "#1c61b6AA")

# Add legend
legend("bottomright",
       legend = c(paste("AUC =", round(auc(roc_obj), 3))),
       col = "blue",
       lwd = 2)

# Calculate and print optimal threshold
optimal_threshold <- coords(roc_obj, "best", ret = "threshold")
print(paste("Optimal threshold:", round(optimal_threshold, 3)))
```



# Task Four: Deploy the prediction model

```{r}
# 1. Create a deployment-ready model interface
deploy_stroke_model <- function() {
  # Load required libraries
  library(caret)
  library(randomForest)
  
  # Create a prediction function
  predict_stroke_risk <- function(new_data) {
    # Load the saved model
    model <- readRDS("stroke_prediction_model.rds")
    
    # Ensure input data has correct format
    required_columns <- c("age", "gender", "hypertension", "heart_disease", 
                         "ever_married", "work_type", "Residence_type", 
                         "avg_glucose_level", "bmi", "smoking_status")
    
    # Convert factors to correct levels
    new_data$gender <- as.factor(new_data$gender)
    new_data$ever_married <- as.factor(new_data$ever_married)
    new_data$work_type <- as.factor(new_data$work_type)
    new_data$Residence_type <- as.factor(new_data$Residence_type)
    new_data$smoking_status <- as.factor(new_data$smoking_status)
    
    # Make prediction
    pred_prob <- predict(model, new_data, type = "prob")
    
    # Return risk score and classification
    risk_score <- pred_prob[, "Stroke"]
    classification <- ifelse(risk_score > 0.5, "High Risk", "Low Risk")
    
    return(list(
      risk_score = risk_score,
      classification = classification,
      probability = pred_prob
    ))
  }
  
  return(predict_stroke_risk)
}

# 2. Example usage function
example_prediction <- function() {
  # Create example patient data
  new_patient <- data.frame(
    age = 65,
    gender = "Male",
    hypertension = 1,
    heart_disease = 0,
    ever_married = "Yes",
    work_type = "Private",
    Residence_type = "Urban",
    avg_glucose_level = 120,
    bmi = 28,
    smoking_status = "formerly smoked"
  )
  
  # Get prediction function
  predict_stroke_risk <- deploy_stroke_model()
  
  # Make prediction
  result <- predict_stroke_risk(new_patient)
  
  # Print results
  cat("Stroke Risk Assessment:\n")
  cat("Risk Score:", round(result$risk_score, 3), "\n")
  cat("Classification:", result$classification, "\n")
  
  return(result)
}

# 3. Save deployment files
saveRDS(rf_model, "stroke_prediction_model.rds")
save(deploy_stroke_model, file = "deploy_functions.RData")

# 4. Create documentation
model_documentation <- "
Stroke Prediction Model Documentation

Input Features Required:
- age: numeric (years)
- gender: factor ('Male' or 'Female')
- hypertension: binary (0 or 1)
- heart_disease: binary (0 or 1)
- ever_married: factor ('Yes' or 'No')
- work_type: factor ('Private', 'Self-employed', 'Govt_job', 'children', 'Never_worked')
- Residence_type: factor ('Urban' or 'Rural')
- avg_glucose_level: numeric (mg/dL)
- bmi: numeric
- smoking_status: factor ('formerly smoked', 'never smoked', 'smokes', 'Unknown')

Output:
- risk_score: probability of stroke (0-1)
- classification: 'High Risk' or 'Low Risk'
- probability: full probability distribution

Model Performance:
- ROC-AUC: [Insert final ROC-AUC score]
- Sensitivity: [Insert final sensitivity]
- Specificity: [Insert final specificity]
"

# Save documentation
writeLines(model_documentation, "model_documentation.txt")

# 5. Test deployment
cat("Testing deployment with example patient...\n")
test_result <- example_prediction()
print(test_result)
```




# Task Five: Findings and Conclusions
































